En Garde! Unguarded Iteration for Reversible Computation in the Delay Monad\thanks{Niccol\`o Veltri was supported by a research grant (13156) from VILLUM FONDEN.}

### 2019-05-03 mpc19/paper.tex
Added figure for iterator.
small fixes

\begin{document}
%
\title{En garde! Unguarded iteration for reversible computation in the
  Delay monad\thanks{Niccol\`o Veltri was supported by a research
    grant (13156) from VILLUM FONDEN.}}
%
\titlerunning{Unguarded iteration for reversible computation in the Delay monad}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Robin Kaarsgaard \inst{1}\orcidID{0000-0002-7672-799X} \and
Niccol\`o Veltri\inst{2}\orcidID{0000-0002-7230-3436}}
%
\authorrunning{Kaarsgaard and Veltri}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{DIKU, Department of Computer Science, University of
    Copenhagen\\ \email{robin@di.ku.dk} \and
Department of Computer Science, IT University of
    Copenhagen\\ \email{nive@itu.dk}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Reversible computation studies computations which exhibit both forward and
backward determinism. Among others, it has been studied for half a century for
its applications in low-power computing, and forms the basis for quantum
computing.

@@

In this paper, we extend the two-level calculus of finitary \Pio{} to one for
full \Pio{} (i.e., with both recursive types and iteration by means of a
\trace{} combinator) using the delay monad, which can be regarded as a
``computability-aware'' analogue of the usual maybe monad for
partiality. This yields a calculus of iterative (and possibly non-terminating)
reversible programs acting on user-defined dynamic data structures together
with a calculus of certified program equivalences between these programs.

@@

The full \Pio\ calculus cannot be modelled in the same framework of
\cite{CaretteS16}, since Martin-L\"of type theory is a total language
which in particolar disallow the specification of a trace
operator. Consequently, it is necessary to move to a setting
supporting the existence of partial maps, and in type theory this can
be done by using monads, by considering partiality as an effect. Our
choice fell on the coinductive delay monad, introduced by Capretta
\cite{Capretta05} as a way of representing general recursive functions
in Martin-L\"of type theory. The delay datatype has been employed in a
large number of applications, ranging from operational semantics of
functional languages \cite{Danielsson12} to formalization of domain
theory in type theory \cite{BentonKV09} and normalization by
evaluation \cite{AbelC14}. Here it is used for giving denotational
semantics to \Pio. In particular, we show how to endow the delay
datatype with a trace combinator, whose construction factors
through the specification of a uniform iteration operator
\cite{GoncharovSRJ18,GoncharovMR16}.

@@

Our work is settled in Martin-L\"of type theory with inductive and
coinductive types. We write $(a : A) \to B \,a$ for dependent function
spaces and $(a : A) \times B \,a$ for dependent products. We allow
dependent functions to have implicit arguments and indicate implicit
argument positions with curly brackets (as in Agda). We use the
symbol $=$ for definitional equality of terms and $\equiv$ for
propositional equality. Given $f : A \to C$ and
$g : B \to C$, we write $\copair f g : A + B \to C$ for their
copairing. The coproduct injections are denoted $\inl$ and $\inr$.
Given $h : C \to A$ and $k : C \to B$, we write
$\pair h k : C \to A \times B$ for their pairing. The product
projections are denoted $\fst$ and $\snd$. The empty type is 0 and the
empty type is 1. We write $\Set$ for the category of
types and functions between them.
We define $A \leftrightarrow B = (A \to B) \times (B \to A)$.
%We write $\U$ for the universe of (small) types.

@@

The types of \Pio\ are given by the grammar:
\[
A ::= \Z \, | \,A \oplus A \, | \,\I \,| \,A \otimes A \,| \,X \,|
\,\mu X.A 
\]
where $X$ ranges over a set of variables. In Agda, we use de Bruijn
indexes to deal with type variables, so the grammar above is formally
realized by the rules in Figure \ref{fig:types}. The type $\Ty\,n$
represents \Pio\ types containing at most $n$ free
variables. Variables themselves are encoded as elements of
$\mathsf{Fin}\,n$, the type of natural numbers strictly smaller then
$n$. Notice that the $\mu$ binds a variable, which, for $A : \Ty\,(n +
1)$, we consider to be $n + 1$.

@@

A selection of term equivalences of \Pio\ is given in Figure
\ref{fig:programequivs}. We only include the equivalences that either
differ or have not previously considered by Carette and Sabry in their
formalization of the finite fragment of \Pio~\cite{CaretteS16}. In
particular, we leave out the long list of Laplaza's coherence axioms
expressing that the types of \Pio\ form a rig category \cite{Laplaza}.
We also omit the equivalences stating that $\lambda_\oplus^{-1}$ is
the total inverse of $\lambda_\oplus$, similarly for the other
structural morphisms.

@@

The delay datatype allows us to program with partial functions, but
the introduced notion of partiality is intensional, in the sense that
computations terminating with the same value in a different number of
steps are considered different. To obtain an extensional notion of
partiality, which in particular allows the specification of a
well-behaved trace operator, we introduce the notion of termination-sensitive
weak bisimilarity.

@@

As an alternative to quotienting the homsets of $\D$, we could have
quotiented the delay datatype by weak bisimilarity:
$\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$. In previous work
\cite{ChapmanUV19}, we showed that this construction has problematic
consequences if we employ Hofmann's approach to quotient types
\cite{Hofmann}. For example, it does not seem possible to lift the
monad structure of $\Delay$ to $\mathsf{Delay}_{\approx}$ without
postulating additional principles such as the axiom of countable
choice. More fundamentally for this work, countable choice would be
needed for modelling the trace operator of \Pio\ in the Kleisli
category of $\mathsf{Delay}_{\approx}$. Notice that, if the setoid approach
to quotienting is employed, the latter constructions go through
without the need for additional assumptions. In order to keep an
agnostic perspective on quotient types and avoid the need for
disputable semi-classical choice principles, we decided to quotient
the homsets of $\D$ by (pointwise) weak bisimilarity instead of the
delay datatype.

@@

Colimits in $\Dapprox$ are inherited from $\Set$. This means that 0 is
also the initial object of $\Dapprox$, similarly $A + B$ is the
binary coproduct of $A$ and $B$ in $\Dapprox$. Given $f : A \to
\Delay C$ and $g : B \to \Delay C$, we define their copairing as
$\copairD f g = \copair f g : A + B \to \Delay C$. The operation
$\copairD - -$ is compatible with weak bisimilarity, in the sense that
$\copairD{f_1}{g_1} \approx \copairD{f_2}{g_2}$ whenever $f_1 \approx f_2$
and $g_1 \approx g_2$.
The coproduct
injections are given by $\inlD = \now \comp \inl : A \to \Delay (A +
B)$ and $\inrD = \now \comp \inr : B \to \Delay (A + B)$.

Just as limits in $\Set$ do not lift to limits in the category $\Par$ of sets
and partial functions, they do not lift to $\Dapprox$ either. This is not an
issue with these concrete formulations of partiality, but rather with the
interaction of partiality (in the sense of restriction categories) and limits
in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is not
terminal object and $A \times B$ is not the binary product of $A$ and $B$ in
$\Dapprox$. In fact, $0$ is (also) the terminal object, with $\lambda \_.\,
\never : A \to \Delay 0$ as the terminal morphism. 
%The binary product of $A$ and $B$ in $\Dapprox$ is given by $A + B + A \times B$. 
Nevertheless, it is possible to prove that 1 and
$\times$ are partial terminal object and partial binary products
respectively, in the sense of Cockett and Lack's restriction
categories \cite{CockettL02,CockettL07}. Here we refrain from making the latter
statement formal. We only show the construction of the partial pairing
operation, which we employ in the interpretation of \Pio.  Given
$f : C \to \Delay A$ and $g : C \to \Delay B$, we define:
\begin{align*}
& \pairD f g : C \to \Delay (A \times B) \\
& \pairD f g = \costr \diamond (\str \comp \pair f g)
\end{align*}
Since the delay monad is commutative, the function $\pairD f g$ is
equal to $\str \diamond (\costr \comp \pair f g)$. The operation
$\pairD - -$ is compatible with weak bisimilarity.

@@

The delay monad is an Elgot monad for which the axioms holds up to
weak bisimilarity, not propositional equality. Its iteration operator
relies on the definition of an auxiliary function $\iterD'$
corecursively defined as follows:
\[ %\arraycolsep=8pt
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\iterD' : (A \to \Delay (B + A)) \to \Delay (B + A) \to \Delay B} \\
\iterD' \,f \,(\now \,(\inl\, b)) &= \now \,b \\
\iterD' \,f \,(\now \,(\inr\, a)) &= \later \,(\iterD' \,f \,(f \,a)) \\
\iterD' \,f \,(\later \,x) &= \later \,(\iterD' \,f \,x) \\
 \\
\multicolumn{2}{l}{\iterD : (A \to \Delay (B + A)) \to A \to \Delay B} \\
\multicolumn{2}{l}{\iterD \,f \,a = \iterD' \,f \,(f \,a)}
\end{array}
\]

@@

As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
up to weak bisimilarity. Here we show the proof of the fixpoint
axiom. The proof relies on an auxiliary proof $\unfolding'$. 
\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (B + A)) \to
  \bind\, \copairD {\now}{\iterD \,f} \approx \iterD' \,f} \\
\unfolding' \,f \,(\now \,(\inl \,b)) & = \now_\approx \\
\unfolding' \,f \,(\now \,(\inr \,a)) & = \laterR_{\approx} \,
  \refl_\approx \\
\unfolding' \,f \,(\later \,x) & = \later_\approx \,(\unfolding'\, f\,
  x) \\
 \\
\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (B + A)) \to
  \copairD {\now}{\iterD \,f}\diamond f \approx \iterD \,f}\\
\multicolumn{2}{l}{\unfolding \,f\,x = \unfolding'\,f\,(f\,x)}
\end{array}
\]

@@

The trace operator in $\Dapprox$ is defined as follows:
\begin{align*}
& \traceD : (A + C \to \Delay (B + C)) \to A \to \Delay B \\
& \traceD \,f = \copairD{\now}{\iterD \,\mapR f}\diamond \mapL f
\end{align*}
The operation $\traceD$ is compatible with weak bisimilarity. 
Pictorially:

\ctikzfig{trace}

Intuitively, the function $\mapL f$ initialises the loop. It either
diverges, so that the trace of $f$ diverges as well, or it terminates. It either terminates with
an element $b : B$, in which case the loop ends immediately returning
$b$, or it converges to a value $c : C$, and in this case we proceed
by  invoking the iteration of $\mapR f$ on $c$.

@@

We conclude this section by remarking that the construction of a trace
operator in the Kleisli category of the maybe monad is impossible
without the assumption of additional classical principles. In fact,
given a map $f : A + C \to B + C + 1$, let $xs$ be the possibly
infinite sequence of elements of $B + C + 1$ produced by the iteration
of $f$ on a given input $A + C$. In order to construct the trace of
$f$, we need to decide whether $xs$ is a finite sequence terminating
with an element of $B + 1$, or $xs$ is an infinite stream of elements
of $C$. This decision requires the limited principle of omniscience,
an instance of the law of excluded middle not provable in Martin-L\"of
type theory:
\[
\mathsf{LPO} = (s : \mathbb{N} \to 2) \to ((n : \mathbb{N}) \times
s\,n \equiv \mathsf{true}) + ((n : \mathbb{N}) \to s\,n \equiv \mathsf{false})
\]
where $2$ is the type of booleans, with $\mathsf{true}$ and
$\mathsf{false}$ as only inhabitants.

@@

Showing that $\traceD$ is a dagger trace operator requires the
construction of an inhabitant of $\traceD\,f\,a \dn b \leftrightarrow
\traceD\,(\daggerD\,f)\,b \dn a$. Thanks to the logical equivalence in
(\ref{eq:traceorbit}), this is equivalent to prove the following
statement instead: 
\[
\Orb f {(\inl\,a)} {(\inl\,b)} \leftrightarrow \Orb {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}
\]
We give a proof of the left-to-right direction, the other implication
is proved in an analogous way.  Notice that a term
$p : \Orb f {(\inl\,a)} {(\inl\,b)}$ can be seen as a finite sequence
of elements of $C$, precisely the intermediate values produced by
$\traceD\,f\,a$ before converging to $b$. The orbit of $b$ wrt. the
partial inverse of $f$ can therefore be computed by reversing the
sequence of elements present in $p$. The construction of the reverse
of an orbit is very similar to the way the reverse of a list is
typically defined in a functional programming language like Haskell.
We first consider an intermediate value $c : C$ and we assume to have
already reversed the initial section of the orbit between $\inl\,a$
and $\inr\,c$, that is a term $p' : \Orb  {(\daggerD\,f)} {(\inr\,c)}
{(\inl\,a)}$. 
\[
\def\arraystretch{1.2}
\begin{array}{lll}
\multicolumn{3}{l}{\reverseOrbit' : (i : \{a : A\} \{b : B\} \to f \,a
\dn b \to \daggerD\,f  \,b \dn a) \to } \\
\multicolumn{3}{l}{\qquad \{a : A\} \{b : B\} \{c : C\} \to } \\
\multicolumn{3}{l}{\qquad \Orb f {(\inr\,c)}
 {(\inl\,b)} \to \Orb  {(\daggerD\,f)} {(\inr\,c)} {(\inl\,a)} \to } \\
\multicolumn{3}{l}{\qquad \Orb  {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}}\\
\reverseOrbit' \,i\, (\done \,p) & p'&= \next\, (i \,p)\,p' \\
\reverseOrbit' \,i\,(\next\,p\,q) & p'&= \reverseOrbit' \,i\,q \,(\next \,(i\,p)\,p')
\end{array}
\]
The proof of $\reverseOrbit'$ proceeds by structural induction on the final
segment of the orbit between $\inr\,c$ and $\inl\,b$ that still needs
to be reversed, which is the argument of type
$\Orb f {(\inr\,c)} {(\inl\,b)}$. There are two possibilities.
\begin{itemize}
\item We have $p : f\,(\inr\,c) \dn
\inl\,b$, in which case $i\,p : \daggerD\,f\,(\inl\,b) \dn
\inl\,c$. Then we return $\next\,(i\,p)\,p'$.
\item There exists another value $c' : C$ such that $p :
  f\,(\inr\,c) \dn \inr\,c'$ and $q : \Orb f {(\inr\,c')}
  {(\inl\,b)}$. Then we recursively invoke the function
  $\reverseOrbit' \,i$ on arguments $q$ and $\next\,(i\,p)\,p' : \Orb
  {(\daggerD \,f)} {(\inr\,c')} {(\inl\,b)}$.
\end{itemize}

The reverse of an orbit is derivable using the auxiliary function $\reverseOrbit'$.
\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\reverseOrbit : (i : \{a : A\} \{b : B\} \to f \,a
  \dn b \to \daggerD\,f  \,b \dn a) \to} \\
\multicolumn{2}{l}{\qquad \{a : A\} \{b : B\} \to} \\
\multicolumn{2}{l}{\qquad \Orb f {(\inl\,a)}
  {(\inl\,b)} \to \Orb  {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}}\\
\reverseOrbit \,i\, (\done \,p) &= \done \, (i \,p) \\
\reverseOrbit \,i\,(\next\,p\,q) &= \reverseOrbit' \,i\,q \,(\done \,(i\,p))
\end{array}
\]
The proof of $\reverseOrbit$ proceeds bystructural induction on the
orbit of type
$\Orb f {(\inl\,a)} {(\inl\,b)}$. There are two possibilities.
\begin{itemize}
\item We have $p : f\,(\inl\,a) \dn
\inl\,b$, in which case $i\,p : \daggerD\,f\,(\inl\,b) \dn
\inl\,a$. Then we return $\done\,(i\,p)$.
\item There exists a value $c : C$ such that $p :
  f\,(\inl\,a) \dn \inr\,c$ and $q : \Orb f {(\inr\,c)}
  {(\inl\,b)}$. We conclude by invoking the function
  $\reverseOrbit' \,i$ on arguments $q$ and $\done\,(i\,p) : \Orb
  {(\daggerD \,f)} {(\inr\,c)} {(\inl\,a)}$.
\end{itemize}

@@

By abuse of notation, we use here $\cdot \times \cdot$ (respectively $\cdot +
\cdot$) to refer to the product (respectively coproduct) in $\Dapprox$ even
though it fails to be a product (respectively coproduct) in $\Inv\Dapprox$.
However, both of these are symmetric monoidal products in $\Inv\Dapprox$, so
their use as objects of $\Inv\Dapprox$ in the interpretation above is justified.

>>>>

\begin{document}
%
\title{En Garde! Unguarded Iteration for Reversible Computation in the
  Delay Monad\thanks{Niccol\`o Veltri was supported by a research
    grant (13156) from VILLUM FONDEN.}}
%
\titlerunning{Unguarded iteration for reversible computation in the Delay monad}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Robin Kaarsgaard \inst{1}\orcidID{0000-0002-7672-799X} \and
Niccol\`o Veltri\inst{2}\orcidID{0000-0002-7230-3436}}
%
\authorrunning{Kaarsgaard and Veltri}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{DIKU, Department of Computer Science, University of
    Copenhagen\\ \email{robin@di.ku.dk} \and
Department of Computer Science, IT University of
    Copenhagen\\ \email{nive@itu.dk}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Reversible computation studies computations which exhibit both forward and
backward determinism. Among others, it has been studied for half a century for
its applications in low-power computing, and forms the basis for quantum
computing.

@@

In this paper, we extend the two-level calculus of finitary \Pio{} to one for
full \Pio{} (i.e., with both recursive types and iteration by means of a
trace combinator) using the delay monad, which can be regarded as a
``computability-aware'' analogue of the usual maybe monad for
partiality. This yields a calculus of iterative (and possibly non-terminating)
reversible programs acting on user-defined dynamic data structures together
with a calculus of certified program equivalences between these programs.

@@

The full \Pio\ calculus cannot be modelled in the same framework of
\cite{CaretteS16}, since Martin-L\"of type theory is a total language
which in particolar disallows the specification of a trace
operator on types. Consequently, it is necessary to move to a setting
supporting the existence of partial maps, and in type theory this can
be done by using monads, by considering partiality as an effect. Our
choice fell on the coinductive delay monad, introduced by Capretta
\cite{Capretta05} as a way of representing general recursive functions
in Martin-L\"of type theory. The delay datatype has been employed in a
large number of applications, ranging from operational semantics of
functional languages \cite{Danielsson12} to formalization of domain
theory in type theory \cite{BentonKV09} and normalization by
evaluation \cite{AbelC14}. Here it is used for giving denotational
semantics to \Pio. In particular, we show how to endow the delay
datatype with a trace combinator, whose construction factors
through the specification of a uniform iteration operator
\cite{GoncharovSRJ18,GoncharovMR16}.

@@

Our work is settled in Martin-L\"of type theory with inductive and
coinductive types. We write $(a : A) \to B \,a$ for dependent function
spaces and $(a : A) \times B \,a$ for dependent products. We allow
dependent functions to have implicit arguments and indicate implicit
argument positions with curly brackets (as in Agda). We use the
symbol $=$ for definitional equality of terms and $\equiv$ for
propositional equality. Given $f : A \to C$ and
$g : B \to C$, we write $\copair f g : A + B \to C$ for their
copairing. The coproduct injections are denoted $\inl$ and $\inr$.
Given $h : C \to A$ and $k : C \to B$, we write
$\pair h k : C \to A \times B$ for their pairing. The product
projections are denoted $\fst$ and $\snd$. The empty type is 0 and the
unit type is 1. We write $\Set$ for the category of
types and functions between them. We also use $\Set$ to denote the
universe of types.
We define $A \leftrightarrow B = (A \to B) \times (B \to A)$.
%We write $\U$ for the universe of (small) types.

@@

The types of \Pio\ are given by the grammar:
\[
A ::= \Z \, | \,A \oplus A \, | \,\I \,| \,A \otimes A \,| \,X \,|
\,\mu X.A 
\]
where $X$ ranges over a set of variables. In Agda, we use de Bruijn
indexes to deal with type variables, so the grammar above is formally
realized by the rules in Figure \ref{fig:types}. The type $\Ty\,n$
represents \Pio\ types containing at most $n$ free
variables. Variables themselves are encoded as elements of
$\mathsf{Fin}\,n$, the type of natural numbers strictly smaller then
$n$. The type constructor $\mu$ binds a variable, which, for $A : \Ty\,(n +
1)$, we consider to be $n + 1$.

@@

A selection of term equivalences of \Pio\ is given in Figure
\ref{fig:programequivs}. We only include the equivalences that either
differ or have not previously considered by Carette and Sabry in their
formalization of the finite fragment of \Pio~\cite{CaretteS16}. In
particular, we leave out the long list of Laplaza's coherence axioms
expressing that types and terms of \Pio\ form a rig category \cite{Laplaza}.
We also omit the equivalences stating that $\lambda_\oplus^{-1}$ is
the total inverse of $\lambda_\oplus$, similarly for the other
structural morphisms.

@@

The delay datatype allows us to program with partial functions, but
the introduced notion of partiality is intensional, in the sense that
computations terminating with the same value in a different number of
steps are considered different. To obtain an extensional notion of
partiality, which in particular allows the specification of a
well-behaved trace operator, we introduce the notion of (termination-sensitive)
weak bisimilarity.

@@

As an alternative to quotienting the homsets of $\D$, we could have
quotiented the delay datatype by weak bisimilarity:
$\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$. In previous work
\cite{ChapmanUV19}, we showed that this construction has problematic
consequences if we employ Hofmann's approach to quotient types
\cite{Hofmann}. For example, it does not seem possible to lift the
monad structure of $\Delay$ to $\mathsf{Delay}_{\approx}$ without
postulating additional principles such as the axiom of countable
choice. More fundamentally for this work, countable choice would be
needed for modelling the trace operator of \Pio\ in the Kleisli
category of $\mathsf{Delay}_{\approx}$. Notice that, if the setoid approach
to quotienting is employed, the latter constructions go through
without the need for additional assumptions. In order to keep an
agnostic perspective on quotient types and avoid the need for
disputable semi-classical choice principles, we decided to quotient
the homsets of $\D$ by (pointwise) weak bisimilarity instead of
the objects of $\D$.

@@

Colimits in $\Dapprox$ are inherited from $\Set$. This means that 0 is
also the initial object of $\Dapprox$, similarly $A + B$ is the
binary coproduct of $A$ and $B$ in $\Dapprox$. Given $f : A \to
\Delay C$ and $g : B \to \Delay C$, their copairing is
$\copairD f g = \copair f g : A + B \to \Delay C$. The operation
$\copairD - -$ is compatible with weak bisimilarity, in the sense that
$\copairD{f_1}{g_1} \approx \copairD{f_2}{g_2}$ whenever $f_1 \approx f_2$
and $g_1 \approx g_2$.
The coproduct
injections are given by $\inlD = \now \comp \inl : A \to \Delay (A +
B)$ and $\inrD = \now \comp \inr : B \to \Delay (A + B)$.

Just as limits in $\Set$ do not lift to limits in the category $\Par$ of sets
and partial functions, they do not lift to $\Dapprox$ either. This is not an
issue with these concrete formulations of partiality, but rather with the
interaction of partiality (in the sense of restriction categories) and limits
in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is
not the
terminal object and $A \times B$ is not the binary product of $A$ and $B$ in
$\Dapprox$. In fact, $0$ is (also) the terminal object, with $\lambda \_.\,
\never : A \to \Delay 0$ as the terminal morphism. 
%The binary product of $A$ and $B$ in $\Dapprox$ is given by $A + B + A \times B$. 
Nevertheless, it is possible to prove that 1 and
$\times$ are partial terminal object and partial binary products
respectively, in the sense of Cockett and Lack's restriction
categories \cite{CockettL02,CockettL07}. Here we refrain from making the latter
statement formal. We only show the construction of the partial pairing
operation, which we employ in the interpretation of \Pio.  Given
$f : C \to \Delay A$ and $g : C \to \Delay B$, we define:
\begin{align*}
& \pairD f g : C \to \Delay (A \times B) \\
& \pairD f g = \costr \diamond (\str \comp \pair f g)
\end{align*}
Since the delay monad is commutative, the function $\pairD f g$ is
equal to $\str \diamond (\costr \comp \pair f g)$. The operation
$\pairD - -$ is compatible with weak bisimilarity.

@@

The delay monad is an Elgot monad for which the axioms holds up to
weak bisimilarity, not propositional equality. In other words, the
category $\Dapprox$ can be endowed with a uniform iteration
operator. The specification of the iteration operator relies on an
auxiliary function $\iterD'$ corecursively defined as follows:
\[ %\arraycolsep=8pt
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\iterD' : (A \to \Delay (B + A)) \to \Delay (B + A) \to \Delay B} \\
\iterD' \,f \,(\now \,(\inl\, b)) &= \now \,b \\
\iterD' \,f \,(\now \,(\inr\, a)) &= \later \,(\iterD' \,f \,(f \,a)) \\
\iterD' \,f \,(\later \,x) &= \later \,(\iterD' \,f \,x) \\
 \\
\multicolumn{2}{l}{\iterD : (A \to \Delay (B + A)) \to A \to \Delay B} \\
\multicolumn{2}{l}{\iterD \,f \,a = \iterD' \,f \,(f \,a)}
\end{array}
\]

@@

As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
up to weak bisimilarity. Here we show the proof of the fixpoint
axiom, which in turns relies on an auxiliary proof $\unfolding'$. 
\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (B + A)) \to
  \bind\, \copairD {\now}{\iterD \,f} \approx \iterD' \,f} \\
\unfolding' \,f \,(\now \,(\inl \,b)) & = \now_\approx \\
\unfolding' \,f \,(\now \,(\inr \,a)) & = \laterR_{\approx} \,
  \refl_\approx \\
\unfolding' \,f \,(\later \,x) & = \later_\approx \,(\unfolding'\, f\,
  x) \\
 \\
\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (B + A)) \to
  \copairD {\now}{\iterD \,f}\diamond f \approx \iterD \,f}\\
\multicolumn{2}{l}{\unfolding \,f\,x = \unfolding'\,f\,(f\,x)}
\end{array}
\]

@@

The trace operator in $\Dapprox$ is defined in terms of the iterator as follows:
\begin{align*}
& \traceD : (A + C \to \Delay (B + C)) \to A \to \Delay B \\
& \traceD \,f = \copairD{\now}{\iterD \,\mapR f}\diamond \mapL f
\end{align*}
The operation $\traceD$ is compatible with weak bisimilarity. Graphically, we express the iterator on $f$ as a wire looping back on the input, i.e., as

\ctikzfig{iter}
In this way, the definition of $\traceD$ may be expressed graphically as

\ctikzfig{trace}
Intuitively, the function $\mapL f$ initialises the loop. It either
diverges, so that the trace of $f$ diverges as well, or it terminates. It either terminates with
an element $b : B$, in which case the loop ends immediately returning
$b$, or it converges to a value $c : C$, and in this case we proceed
by  invoking the iteration of $\mapR f$ on $c$.

@@

We conclude this section by remarking that the construction of a trace
operator in the Kleisli category of the maybe monad is impossible
without the assumption of additional classical principles. In fact,
given a map $f : A + C \to B + C + 1$, let $xs$ be the possibly
infinite sequence of elements of $B + C + 1$ produced by the iteration
of $f$ on a given input in $A$. In order to construct the trace of
$f$, we need to decide whether $xs$ is a finite sequence terminating
with an element of $B + 1$, or $xs$ is an infinite stream of elements
of $C$. This decision requires the limited principle of omniscience,
an instance of the law of excluded middle not provable in Martin-L\"of
type theory:
\[
\mathsf{LPO} = (s : \mathbb{N} \to 2) \to ((n : \mathbb{N}) \times
s\,n \equiv \mathsf{true}) + ((n : \mathbb{N}) \to s\,n \equiv \mathsf{false})
\]
where $2$ is the type of booleans, with $\mathsf{true}$ and
$\mathsf{false}$ as only inhabitants.

@@

Showing that $\traceD$ is a dagger trace operator requires the
construction of an inhabitant of $\traceD\,f\,a \dn b \leftrightarrow
\traceD\,(\daggerD\,f)\,b \dn a$. Thanks to the logical equivalence in
(\ref{eq:traceorbit}), this is equivalent to prove the following
statement instead: 
\[
\Orb f {(\inl\,a)} {(\inl\,b)} \leftrightarrow \Orb {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}
\]
We give a detailed proof of the left-to-right direction, the other implication
is derived  in an analogous way.  Notice that a term
$p : \Orb f {(\inl\,a)} {(\inl\,b)}$ can be seen as a finite sequence
of elements of $C$, precisely the intermediate values produced by
$\traceD\,f\,a$ before converging to $b$. The orbit of $b$ wrt. the
partial inverse of $f$ can therefore be computed by reversing the
sequence of elements present in $p$. The construction of the reverse
of an orbit is very similar to the way the reverse of a list is
typically defined in a functional programming language like Haskell.
We first consider an intermediate value $c : C$ and we assume to have
already reversed the initial section of the orbit between $\inl\,a$
and $\inr\,c$, that is a term $p' : \Orb  {(\daggerD\,f)} {(\inr\,c)}
{(\inl\,a)}$. 
\[
\def\arraystretch{1.2}
\begin{array}{lll}
\multicolumn{3}{l}{\reverseOrbit' : (i : \{a : A\} \{b : B\} \to f \,a
\dn b \to \daggerD\,f  \,b \dn a) \to } \\
\multicolumn{3}{l}{\qquad \{a : A\} \{b : B\} \{c : C\} \to } \\
\multicolumn{3}{l}{\qquad \Orb f {(\inr\,c)}
 {(\inl\,b)} \to \Orb  {(\daggerD\,f)} {(\inr\,c)} {(\inl\,a)} \to } \\
\multicolumn{3}{l}{\qquad \Orb  {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}}\\
\reverseOrbit' \,i\, (\done \,p) & p'&= \next\, (i \,p)\,p' \\
\reverseOrbit' \,i\,(\next\,p\,q) & p'&= \reverseOrbit' \,i\,q \,(\next \,(i\,p)\,p')
\end{array}
\]
The proof of $\reverseOrbit'$ proceeds by structural induction on the final
segment of the orbit between $\inr\,c$ and $\inl\,b$ that still needs
to be reversed, which is the argument of type
$\Orb f {(\inr\,c)} {(\inl\,b)}$. There are two possibilities.
\begin{itemize}
\item We have $p : f\,(\inr\,c) \dn
\inl\,b$, in which case $i\,p : \daggerD\,f\,(\inl\,b) \dn
\inl\,c$. Then we return $\next\,(i\,p)\,p'$.
\item There exists another value $c' : C$ such that $p :
  f\,(\inr\,c) \dn \inr\,c'$ and $q : \Orb f {(\inr\,c')}
  {(\inl\,b)}$. Then we recursively invoke the function
  $\reverseOrbit' \,i$ on arguments $q$ and $\next\,(i\,p)\,p' : \Orb
  {(\daggerD \,f)} {(\inr\,c')} {(\inl\,b)}$.
\end{itemize}

The reverse of an orbit is derivable using the auxiliary function $\reverseOrbit'$.
\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\reverseOrbit : (i : \{a : A\} \{b : B\} \to f \,a
  \dn b \to \daggerD\,f  \,b \dn a) \to} \\
\multicolumn{2}{l}{\qquad \{a : A\} \{b : B\} \to} \\
\multicolumn{2}{l}{\qquad \Orb f {(\inl\,a)}
  {(\inl\,b)} \to \Orb  {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}}\\
\reverseOrbit \,i\, (\done \,p) &= \done \, (i \,p) \\
\reverseOrbit \,i\,(\next\,p\,q) &= \reverseOrbit' \,i\,q \,(\done \,(i\,p))
\end{array}
\]
The proof of $\reverseOrbit$ proceeds by structural induction on the
orbit of type
$\Orb f {(\inl\,a)} {(\inl\,b)}$. There are two possibilities.
\begin{itemize}
\item We have $p : f\,(\inl\,a) \dn
\inl\,b$, in which case $i\,p : \daggerD\,f\,(\inl\,b) \dn
\inl\,a$. Then we return $\done\,(i\,p)$.
\item There exists a value $c : C$ such that $p :
  f\,(\inl\,a) \dn \inr\,c$ and $q : \Orb f {(\inr\,c)}
  {(\inl\,b)}$. We conclude by invoking the function
  $\reverseOrbit' \,i$ on arguments $q$ and $\done\,(i\,p) : \Orb
  {(\daggerD \,f)} {(\inr\,c)} {(\inl\,a)}$.
\end{itemize}

@@

By abuse of notation, we use here $\times$ (respectively $+$) to refer
to the product (respectively coproduct) in $\Dapprox$ even though it
fails to be a product (respectively coproduct) in $\Inv\Dapprox$.
However, both of these are symmetric monoidal products in
$\Inv\Dapprox$, so their use as objects of $\Inv\Dapprox$ in the
interpretation above is justified.
